### 2024-09-26 manuscript.tex
Various improvements

Previous studies have identified substantial large-scale biases in model clouds
over the Southern Ocean (SO), affecting sea surface temperature and the Earth's
albedo overall. Our aim is to quantify how well the high-resolution ICON model
is simulating clouds in this region, particularly in light of the fact that
subgrid-scale clouds are not parametrised in this model. This region is mostly
dominated by boundary layer (BL) clouds generated by shallow convection, and
these are problematic to observe by spaceborne lidar and radars, which are
affected by attenuation by overlapping and thick clouds and ground clutter,
respectively. Therefore, we choose to use a large set of ship-based
observations conducted with ceilometers and lidars on board of the Research
Vessel (RV) \textit{Polarstern} and other voyages. Altogether, we analyse over
2000 days (about 6 years) of data from 31 voyages and 1 sub-antarctic station
covering diverse longitudes of the SO. To achieve a like-for-like comparison
with the model, we use a ground-based lidar simulator called the Automatic
Lidar and Ceilometer Framework (ALCF). We contrast the results with the ECMWF
Reanalysis 5 (ERA5) and the Modern-Era Retrospective analysis for Research and
Applications, Version 2 (MERRA-2).

@@

The main aim of our research is to evaluate the GSRM version of ICON, developed
jointly by nextGEMS, Deutscher Wetterdienst (DWD) and the Max-Planck-Institute
for Meteorology (MPI-M). Previous studies have identified substantial
large-scale biases in climate model clouds over the Southern Ocean (SO),
affecting sea surface temperature and the Earth’s albedo overall. Our aim is to
quantify how well the high-resolution ICON model is simulating clouds in this
region, particularly in light of the fact that subgrid-scale clouds are not
parameterised in this model. This region is mostly dominated by BL clouds
generated by shallow convection, and these are problematic to observe by
spaceborne lidar and radars, which are affected by attenuation by overlapping
and thick clouds and ground clutter, respectively.  Specifically, the radar on
CloudSat and lidar on CALIPSO are affected by the abovementioned issues,
resulting in a strong underestimation of cloud occurrence below 2 km relative
to ground-based lidar observations \citep{mcerlich2021}.  This, in turn, can
lead to systematic biases in low clouds in climate models, which are frequently
evaluated against CloudSat--CALIPSO products. Reanalyses can also suffer from
cloud biases, as these are usually parametrised in their atmospheric component,
and also in regions where input observations are sparse.  This makes them a
problematic reference for clouds over the SO, and any biases relative to a
reanalysis should be interpreted with caution. Instead, we chose to use a large
set of ship-based observations conducted with ceilometers and lidars on board
of the RV \emph{Polarstern} and other voyages and stations as a reference for
the model evaluation.

@@

\subsection{Cloud occurrence by height}

@@

When subsetted by low- and high-latitude zones (Fig.
\ref{fig:cloud-occurrence}b, c), we can see that the low-latitude SO zone shows
a stronger peak of cloud occurrence near the surface than the high-latitude SO
zone, and this could be because higher latitudes have more unstable atmospheric
profiles. The low- and high-latitude SO zones show similar biases in models as
in the general case, but ERA5 does not overestimate the peak in the
low-latitude SO zone (near-surface cloud occurrence is still strongly
understimated).

When subsetted by cyclonic and non-cyclonic situations (Fig.
\ref{fig:cloud-occurrence}d, e), we can see that the cyclonic situations have a
larger amount of observed cloudiness, including the peak and total cloud
fraction. In these situations, the models are doing a relatively good job of
getting the vertical profile of cloud occurrence right, but still tend to
underestimate cloud occurrence above 1 km and near the surface. Non-cyclonic
situations are similar to the general case.

When subsetted by the relatively stable and unstable conditions (Fig.
\ref{fig:cloud-occurrence}f, g), as defined in Section
\ref{sec:cyclone-stability}, we can see that in the relatively stable
situations cloud occurrence peaks strongly at the surface in observations,
compared to the relatively unstable situations, where the peak is more obtuse
and spread across 0--1 km.  In the relatively stable situations, the models are
doing a fairly good job, but overestimate cloud occurrence at the peak below 1
km; above 1 km, they show similar biases as in the general case.  In the
relatively unstable situations the bias in ICON is very pronounced, with a much
stronger peak at about 500 m, ERA5 is underestimating cloud occurrence below 1
km (especially near the surface), and MERRA-2 even more strongly.

@@

\begin{figure}[p!]
\centering
\includegraphics[width=\textwidth]{img/cl_agg.pdf}
\caption{
Cloud occurrence by height calculated as the average of all voyages and
stations for the observed (OBS) and simulated lidar cloud mask. The total cloud
fraction (CF), average shortwave (SW) and longwave (LW) are shown in the
legend, and the relative frequency of occurrence (RFO) of the subset is shown
below.  The bands span the 16$^\mathrm{th}$--84$^\mathrm{th}$ percentile
calculated from the set of all voyages and stations. The subsets
\textbf{(a--g)} are described in the text (Section \ref{sec:results}).
}
\label{fig:cloud-occurrence}
\end{figure}

@@

We analysed the daily cloud cover (total cloud fraction) distribution. This is
a measure of cloudiness, irrespective of height. A cloud detected at any height
means that the lidar profile was classified as cloudy; otherwise, it was
classified as a clear sky. When all profiles in a day are aggregated together,
the cloud cover for the day is defined as the fraction of cloudy profiles in
the total number of profiles, expressed in oktas.

In Fig.  \ref{fig:cloud-cover} we show the results for the same subsets as
analysed above. Observations show the greatest representation of high cloud
cover (5--8 oktas), peaking at 7 oktas. This is not represented by ICON or the
reanalyses.  While ICON is the closest, it tends to be 1 okta clearer the
observations, peaking at 6 oktas, and highly understimating days with 8 oktas.
Overall, the reanalyses show results similar to each other, underestimating
cloud cover by about 2 oktas, and strongly understimating days with 7 and 8
oktas. Of the two reanalyses, MERRA-2 shows slightly higher cloud cover and
thus more consistent with observations.

When analysed by subsets, the cyclonic subset shows the highest cloud cover,
with 8 oktas occurring on one half of such days (Fig. \ref{fig:cloud-cover}d).
This is not represented by ICON or the reanalyses at all. Interestingly, clear
sky days (0 oktas) also have a local maximum peaking at about 15\% in this
subset.  When we contrast the low and high latitude zones, we see that the high
latitude zone tends to have greater cloud cover, peaking at 8 oktas (Fig.
\ref{fig:cloud-cover}c). The high latitude also has almost no clear sky or
small cloud cover cases (0--4 oktas). ICON and the reanalyses represent at
least this characteristic of the distribution well for 0--3 oktas, but
otherwise show bises similar to the general case. One of the greatest biases is
present in ERA5 in the relatively unstable subset, where it peaks at 3 oktas, whereas the
observations peak at 7 oktas and show negligible cloud cover below 5 oktas.

@@

Limitations of our study are numerous. The spatial coverage of our dataset
does not include most parts of the the Indian Ocean and Pacific Ocean sectors
of the SO. Even though climatological features of the SO are typically
relatively uniform zonally, variantions exist, such as those related to the
Antarctic Peninsula and the southern of South America. The voyages were
mostly undertaken in the Austral summer months, and only rarely in the winter
months, due to the poor accesibility of this region during winter. Therefore,
our results are mostly representative of summer, and to lesser extent spring
and autumn conditions.

The time period of ICON is relatively short with only four full years of
simulation available. Moreover, the simulation was free running, which means
that observations had to be temporally mapped to this time period (at the same
time relative to the start of the year) for the comparison. For these reasons,
one can expect the results to be slightly different due to reasons unrelated to
model biases, such as different weather conditions and the phase of climatic
oscillations such as the ENSO in the observations and the model.

Ground-based lidar observations are affected by attenuation by thick cloud
layers, and for this reason the results are mostly representative of BL clouds,
while higher-level clouds are only occassionaly visible to the lidar when BL
are not present. Ground-based lidar observations can be regarded as superior to
satellite lidar observations for low-level clouds in this region, while mid-
and high-level clouds are better represented in satellite observations
\citep{mcerlich2021}.

We attempted to remove lidar profiles with precipitation, which could not be
properly simulated with the lidar simulator (Section \ref{sec:ann}). However,
the approach was limited by relatively low sensitivity of the ANN (65\%), and
the fact that we had to choose a fixed threshold for surface precipitation flux
in the model and renalyses, which might not exactly correspond to detection by
the ANN applied to obseravations. Also, we did not make an attempt to remove
profiles with precipitation not reaching the surface. The above reasons can
result in artificial bias in the comparison, even though we exect this to be
much smaller than the identified model biases.

@@

We analysed a total of about 2421 days of lidar and radiosonde data from 31
voyages and campaigns and one subantarctic station data, covering the Atlantic,
Australian and New Zealand sectors of the SO over the span of 10 years. This
dataset, together with the use of the ALCF lidar simulator, provided a fairly
comprehensive basis for evaluating SO cloud and thermodynamic profile biases in
the GSRM ICON and the ERA5 and MERRA-2 reanalyses. This provides a unique
evaluation perspective different from satellite observations, one which is more
suitable for evaluating bounary layer clouds, which are predominant in this
region. Furthermore, we subsetted our dataset by low and high latitude bands,
cyclonic activity and stability, in order to identify how these conditions
relate the the biases. Our main finding corroborates previous findings of large
BL cloud biases in models, and their subsequent effect on the radiative
transfer. This also applies to the new ICON GSRM, but the biases are lower than
in the reanalyses, despite the reanalyses having the advantage of assimilation
of the observed meteorological conditions.  The GSRM has, on the other hand,
the advange of a much higher spatial resolution and, to some extent, explicit
calculation of traditionally subgrid-scale processes such as convection.

@@

We acknowledge the RV Polarstern datasets provided by the Alfred Wegener
Institute, RSV Aurora Australis datasets provided by the Australian Antarctic
Program and the University of Canterbury (UC), RV Tangaroa datasets provided by
the National Institute of Water and Atmospheric Research and the UC, RV
Nathaniel B. Palmer dataset provided by the National Science Foundation,
Cooperative Institute for Research in Environmental Sciences, University of
Colorado and the UC, HMNZS Wellington dataset provided by the Royal New Zealand
Navy and the UC and a Macquarie Island dataset provided by the Atmospheric
Radiation Measurement and the UC. We thank the scientific staff, the crew, and
everyone involved in collecting data on the voyages and Macquarie Island,
especially Kelly Schick, Peter Guest, John J. Cassano, Simon Parsons, Graeme
Plank, Sally Garrett, Jamie Halla, Sean Hartery, Mike J.  Harvey, Simon P.
Alexander and Alex Schuddeboom. We acknowledge the reanalysis dataset ERA5
provided by the Copernicus Climate Change Service and MERRA-2 provided by the
Global Modeling and Assimilation Office. We also acknowledge the AMSR Unified
and SMMR and SSM/I-SSMIS datasets provided by the National Snow and Ice Data
Center.

@@

The RV \emph{Polarstern} data are openly available from Pangaea
(\url{https://pangaea.de}). The MARCUS and Macquarie Island (MICRE) data are
openly available from the Atmospheric Radiation Measurement (ARM)
(\url{https://www.arm.gov}). The MERRA-2 data are openly available from the
NASA Goddard Earth Sciences (GES) Data and Information Services Center
(DISC) (\url{https://disc.gsfc.nasa.gov/datasets?project=MERRA-2}).  The ERA5
data are openly available from the Copernicus Climate Data Store (CDS)
(\url{https://cds.climate.copernicus.eu}). The ICON data are available on the
Levante cluster of the Deutsches Klimarechenzentrum (DKRZ;
\url{https://www.dkrz.de/en/systems/hpc/hlre-4-levante}) after registration at
\url{https://luv.dkrz.de/register/}. The CERES products are available from the
project website (\url{https://ceres.larc.nasa.gov}) and the NASA Atmospheric
Science Data Centre (\url{https://asdc.larc.nasa.gov/project/CERES}). The other
voyage and station data are openly available on Zenodo (). Subsetted model and
reanalysis data for the voyage tracks and the station and data processed with
the ALCF are openly available on Zenodo ().

>>>>

Previous studies have identified substantial large-scale biases in model clouds
over the Southern Ocean (SO), affecting sea surface temperature and the Earth's
albedo overall. Our aim is to quantify how well the high-resolution ICON model
is simulating clouds in this region, particularly in light of the fact that
subgrid-scale clouds are not parametrised in this model. This region is mostly
dominated by boundary layer clouds generated by shallow convection, and
these are problematic to observe by spaceborne lidar and radars, which are
affected by attenuation by overlapping and thick clouds and ground clutter,
respectively. Therefore, we choose to use a large set of ship-based
observations conducted with ceilometers and lidars on board of the Research
Vessel (RV) \textit{Polarstern} and other voyages. Altogether, we analyse over
2000 days (about 6 years) of data from 31 voyages and 1 sub-antarctic station
covering diverse longitudes of the SO. To achieve a like-for-like comparison
with the model, we use a ground-based lidar simulator called the Automatic
Lidar and Ceilometer Framework (ALCF). We contrast the results with the ECMWF
Reanalysis 5 (ERA5) and the Modern-Era Retrospective analysis for Research and
Applications, Version 2 (MERRA-2).

@@

The main aim of our research is to evaluate the GSRM version of ICON, developed
jointly by nextGEMS, Deutscher Wetterdienst (DWD) and the Max-Planck-Institute
for Meteorology (MPI-M). Previous studies have identified substantial
large-scale biases in climate model clouds over the Southern Ocean (SO),
affecting sea surface temperature and the Earth’s albedo overall. Our aim is to
quantify how well the high-resolution ICON model is simulating clouds in this
region, particularly in light of the fact that subgrid-scale clouds are not
parameterised in this model. This region is mostly dominated by boundary layer
clouds generated by shallow convection, and these are problematic to observe by
spaceborne lidar and radars, which are affected by attenuation by overlapping
and thick clouds and ground clutter, respectively.  Specifically, the radar on
CloudSat and lidar on CALIPSO are affected by the abovementioned issues,
resulting in a strong underestimation of cloud occurrence below 2 km relative
to ground-based lidar observations \citep{mcerlich2021}.  This, in turn, can
lead to systematic biases in low clouds in climate models, which are frequently
evaluated against CloudSat--CALIPSO products. Reanalyses can also suffer from
cloud biases, as these are usually parametrised in their atmospheric component,
and also in regions where input observations are sparse.  This makes them a
problematic reference for clouds over the SO, and any biases relative to a
reanalysis should be interpreted with caution. Instead, we chose to use a large
set of ship-based observations conducted with ceilometers and lidars on board
of the RV \emph{Polarstern} and other voyages and stations as a reference for
the model evaluation.

@@

\subsection{Cloud occurrence by height}
\label{sec:cloud-occurrence}

@@

When subsetted by low- and high-latitude zones (Fig.
\ref{fig:cloud-occurrence}b, c), we see that the low-latitude SO zone shows
a stronger peak of cloud occurrence near the surface than the high-latitude SO
zone, and this could be because higher latitudes have more unstable atmospheric
profiles. The low- and high-latitude SO zones show similar biases in models as
in the general case, but ERA5 does not overestimate the peak in the
low-latitude SO zone (near-surface cloud occurrence is still strongly
understimated).

When subsetted by cyclonic and non-cyclonic situations (Fig.
\ref{fig:cloud-occurrence}d, e), we see that the cyclonic situations have a
larger amount of observed cloudiness, including the peak and total cloud
fraction. In these situations, the models are doing a relatively good job of
getting the vertical profile of cloud occurrence right, but still tend to
underestimate cloud occurrence above 1 km and near the surface. Non-cyclonic
situations are similar to the general case.

When subsetted by the relatively stable and unstable conditions (Fig.
\ref{fig:cloud-occurrence}f, g), as defined in Section
\ref{sec:cyclone-stability}, we see that in the relatively stable
situations cloud occurrence peaks strongly at the surface in observations,
compared to the relatively unstable situations, where the peak is more obtuse
and spread across the altitudes of 0--1 km.  In the relatively stable
situations, the models are doing a fairly good job, but overestimate cloud
occurrence at the peak below 1 km; above 1 km, they show similar biases as in
the general case.  In the relatively unstable situations, the bias in ICON is
very pronounced, with a much stronger peak at about 500 m, ERA5 is
underestimating cloud occurrence below 1 km (especially near the surface), and
MERRA-2 is underestimating it even more strongly.

@@

\begin{figure}[p!]
\centering
\includegraphics[width=\textwidth]{img/cl_agg.pdf}
\caption{
Cloud occurrence by height calculated as the average of all voyages and
stations for the observed (OBS) and simulated lidar cloud mask. The total cloud
fraction (CF), average shortwave (SW) and longwave (LW) are shown in the
legend, and the relative frequency of occurrence (RFO) of the subset is shown
below.  The bands span the 16$^\mathrm{th}$--84$^\mathrm{th}$ percentile
calculated from the set of all voyages and stations. The subsets
\textbf{(d--g)} are defined in Section \ref{sec:cyclone-stability}.
}
\label{fig:cloud-occurrence}
\end{figure}

@@

We analysed the daily cloud cover (total cloud fraction) distribution. This is
a measure of cloudiness, irrespective of height, calculated over the course of
a day (UTC). A cloud detected at any height means that the lidar profile was
classified as cloudy; otherwise, it was classified as a clear sky. When all
profiles in a day are taken together, the cloud cover for the day is defined as
the fraction of cloudy profiles in the total number of profiles, expressed in
oktas (multiples of 1/8).

In Fig.  \ref{fig:cloud-cover} we show the results for the same subsets of data
as in the previous section. Observations show the greatest representation of
high cloud cover (5--8 oktas), peaking at 7 oktas. This is not represented by
ICON or the reanalyses.  While ICON is the closest, it tends to be 1 okta
clearer the observations, peaking at 6 oktas, and highly understimating days
with 8 oktas.  Overall, the reanalyses show results similar to each other,
underestimating cloud cover by about 2 oktas, and strongly understimating days
with 7 and 8 oktas. Of the two reanalyses, MERRA-2 shows slightly higher cloud
cover and thus is more consistent with observations.

When analysed by subsets, the cyclonic subset shows the highest cloud cover,
with 8 oktas occurring on one half of such days (Fig. \ref{fig:cloud-cover}d).
This is not represented by ICON or the reanalyses at all. Interestingly, clear
sky days (0 oktas) also have a local maximum peaking at about 15\% in this
subset.  When we contrast the low- and high-latitude zones, we see that the
high-latitude zone tends to have greater cloud cover, peaking at 8 oktas (Fig.
\ref{fig:cloud-cover}c). The high-latitude zone also has almost no clear sky or
small cloud cover cases (0--4 oktas). ICON and the reanalyses represent at
least this characteristic of the distribution well for 0--3 oktas, but
otherwise show bises similar to the general case. One of the greatest biases is
present in ERA5 in the relatively unstable subset, in which ERA5 peaks at 3
oktas, whereas the observations peak at 7 oktas and show negligible cloud cover
below 5 oktas.

@@

Let us consider the main limitations of the presented results. The spatial
coverage of our dataset does not include most parts of the the Indian Ocean and
Pacific Ocean sectors of the SO. Even though climatological features of the SO
are typically relatively uniform zonally, variantions exist, such as those
related to the Antarctic Peninsula and the southern of South America. The
voyages were mostly undertaken in the Austral summer months, and only rarely in
the winter months, due to the poor accesibility of this region during winter.
Therefore, our results are mostly representative of summer, and to lesser
extent spring and autumn conditions.

The time period of ICON is relatively short with only four full years of
simulation available. Moreover, the simulation is free-running, which means
that observations had to be temporally mapped to this time period (at the same
time relative to the start of the year) for the comparison. For these reasons,
one can expect the results to be slightly different due to reasons unrelated to
model biases, such as different weather conditions and the phase of climate
oscillations such as the ENSO in the observations and the model.

Ground-based lidar observations are affected by attenuation by thick cloud
layers, and for this reason the results are mostly representative of boundary
layer clouds, while higher-level clouds are only occassionaly visible to the
lidar when boundary layer clouds are not present. Ground-based lidar
observations can be regarded as superior to satellite lidar observations for
low-level clouds in this region, while mid- and high-level clouds are better
represented in satellite observations \citep{mcerlich2021}.

We attempted to remove lidar profiles with precipitation, which could not be
properly simulated with the lidar simulator (Section \ref{sec:ann}). However,
the approach was limited by relatively low sensitivity of the ANN (65\%), and
the fact that we had to choose a fixed threshold for surface precipitation flux
in the model and renalyses, which might not exactly correspond to detection by
the ANN applied to obseravations. Also, we did not make an attempt to remove
profiles with precipitation not reaching the surface. The above reasons can
result in an artificial bias in the comparison, even though we expect this to
be much smaller than the identified model biases.

@@

We analysed a total of about 2421 days of lidar and radiosonde data from 31
voyages and campaigns and one subantarctic station data, covering the Atlantic,
Australian and New Zealand sectors of the SO over the span of 10 years. This
dataset, together with the use of the ALCF lidar simulator, provided a fairly
comprehensive basis for evaluating SO cloud and thermodynamic profile biases in
the GSRM ICON and the ERA5 and MERRA-2 reanalyses. This provides a unique
evaluation perspective different from satellite observations, one which is more
suitable for evaluating bounary layer clouds, which are predominant in this
region. Furthermore, we subsetted our dataset by low and high latitude bands,
cyclonic activity and stability, in order to identify how these conditions
relate the the biases. Our main finding corroborates previous findings of large
boundary layer cloud biases in models, and their subsequent effect on the
radiative transfer. This also applies to the new ICON GSRM, but the biases are
lower than in the reanalyses, despite the reanalyses having the advantage of
assimilation of the observed meteorological conditions.  The GSRM has, on the
other hand, the advange of a much higher spatial resolution and, to some
extent, explicit calculation of traditionally subgrid-scale processes such as
convection.

@@

We acknowledge the RV \emph{Polarstern} datasets provided by the Alfred Wegener
Institute and Pangaea, RSV \emph{Aurora Australis} datasets provided by the
Australian Antarctic Program and the University of Canterbury (UC), RV
\emph{Tangaroa} datasets provided by the National Institute of Water and
Atmospheric Research and the UC, RV \emph{Nathaniel B. Palmer} dataset provided
by the National Science Foundation, Cooperative Institute for Research in
Environmental Sciences, University of Colorado and the UC, HMNZS
\emph{Wellington} dataset provided by the Royal New Zealand Navy and the UC and
the Macquarie Island station dataset provided by the Atmospheric Radiation
Measurement and the UC.  We thank the scientific staff, the crew, and everyone
involved in collecting data on the voyages and stations, especially Kelly
Schick, Peter Guest, John J.  Cassano, Simon Parsons, Graeme Plank, Sally
Garrett, Jamie Halla, Sean Hartery, Mike J.  Harvey, Simon P.  Alexander and
Alex Schuddeboom. We acknowledge the ICON model output provided by the
Deutsches Klimarechenzentrum (DKRZ), reanalysis dataset ERA5 provided by the
Copernicus Climate Change Service and MERRA-2 provided by the Global Modeling
and Assimilation Office.

@@

The RV \emph{Polarstern} data are openly available from Pangaea
(\url{https://pangaea.de}). The MARCUS and Macquarie Island (MICRE) data are
openly available from the Atmospheric Radiation Measurement (ARM)
(\url{https://www.arm.gov}). The MERRA-2 data are openly available from the
NASA Goddard Earth Sciences (GES) Data and Information Services Center (DISC)
(\url{https://disc.gsfc.nasa.gov/datasets?project=MERRA-2}).  The ERA5 data are
openly available from the Copernicus Climate Data Store (CDS)
(\url{https://cds.climate.copernicus.eu}). The ICON data are available on the
Levante cluster of the DKRZ
(\url{https://www.dkrz.de/en/systems/hpc/hlre-4-levante}) after registration at
\url{https://luv.dkrz.de/register/}. The CERES products are available from the
project website (\url{https://ceres.larc.nasa.gov}) and the NASA Atmospheric
Science Data Centre (\url{https://asdc.larc.nasa.gov/project/CERES}). The other
voyage and station data are openly available on Zenodo (). Subsetted model and
reanalysis data for the voyage tracks and the station and data processed with
the ALCF are openly available on Zenodo ().
